Attribute fusion in a latent process model for time series of graphs

### 2012-06-22 response.tex

We thank the referees for their feedback. We have
revised the paper in light of their comments and criticism, especially 
on issues regarding the paper appropriateness for the Trans on SP
readership. A more detailed response to the comments from each referee
is included below.

=================
Referee 1

Although the idea of linearly fusing attributes seems appealing and
the authors present their ideas in a technically sound way, nicely
structured and well supported by theoretical analysis and derivations,
I have some critical reservations:

1. The practicality of the approach: Both examples consider the case
of graphs with k=2 attributes and a convenient single parameter
representation ($\theta$). However for larger k, sets of different
optimal values would be harder to argue about. Also given a set of
real graph snapshots the determination (or approximation) of the
corresponding Markov processes - assuming all properties assumed hold
- is not commented on.

*** Response to point 1:
    The cases where $ k \geq 3$ was not considered in the experiments,
    as we deemed them to be more complex but not more
    difficult. Theoretically, the optimal $\lambda$ for $k \geq 3$ can
    be found for all of the graph invariants by numerical techniques,
    i.e., find the $\lambda$ that maximizes non-central parameters
    $\mu_{\lambda}$ in the cases for size and number of triangles, or
    the ratio $\rho_{\lambda}/\zeta_{\lambda}$ in the case of max
    degree or scan. As for empirical evaluation, one can parametrize 
    the set $\{\lambda \colo \| \lambda \| = 1\}$ for $k \geq 3$
    using hypershperical coordinates and the problem of finding the
    optimal $\lambda$ can be done by exhaustive search. This is 
    inefficient or even intractable for $k \geq 5$, but this is a
    common issue for most global optimization problems.

    The issue of determining or estimating the parameters of the
    underlying Markov chain given a set of snapshots of the graphs is an
    important issue that we failed to comment on explicitly in the
    previous submission. We have added in a paragraph to the revision that
    allude to this issue. In general, it is intractable to estimate the
    parameters of the underlying Markov chains given the snapshots. It is
    possible, however, to estimate the latent vectors associated with the
    first and second-order approximation via a spectral embedding
    technique [1]. We also note that more refined models, where the
    messaging events between each pairs of vertices is modeled as point
    process, e.g. [2,3] can also be formulated. The snapshot of the graphs
    is then a binning of these point processes and the issue of parameters
    estimation is fundamental but feasible in these point processes
    approaches.

    [1] Sussman et. al. ``A consistent dot product embedding for
    stochastic blockmodel graphs'', Online at http://arxiv.org/abs/1108.2228

    [2] Heard et. al. ``Bayesian anomaly detection methods for social
    networks'', Annals of Applied Statistics, (4):645--662, 2010.

    [3] P. Perry and P. J. Wolfe. ``Point process modeling for directed
    interaction networks'', Online at http://arxiv.org/abs/1011.1703v2

***

2. Journal relevance: This is very important since there does not seem
to exist some clear identification of a concrete application from
signal processing. The content would be much more relevant to a
statistics journal or to network analysis audience.

*** Response to point 2:

We have added a new section to motivates the consideration of
time-series of attributed graphs. We believe that the notion of a
time-series is a fundamental concept in signal processing and
furthermore the representation of data as (attributed) graphs is
ubiquitous in many applications domain. Also, anomaly/change-point
detection is one of the main problems in time-series analysis. As
such, anomaly detection on time-series of graphs is a natural and
challenging problem. There have been recent work on signal processing
on graphs, with many of them on the construction of wavelets on
graphs. Our paper is similar in spirit to these wavelets paper in that
they all consider signal processing on non-Euclidean data (graphs in
particular), but we have taken a more time-series and statistical 
point of view, as evidenced by our setup of the anomaly detection
problem via hypothesis testings along with the derivations of the
limiting behaviour of the test statistics.  
***

=======
Referee 2

*** Comments
The contributions of this paper are timely, relevant, and novel. The
development of signal processing methods for graphs is an important
topic to the signal processing community, as noted by the special
sessions at recent ICASSP and SSP conferences. The literature on
analysis methods for dynamic graphs is limited, and even more so for
attributed graphs. Furthermore, theoretical results that are also
practically useful are extremely rare.

The main weakness of this paper is that it is not organized well and
is not self-contained.

As such, it is difficult to understand the latent process model
without referring to the previous work (reference [1]). There really
is no introduction to the paper, no context is provided for the
problem--aside from the first sentence of the abstract, the reader is
left with no idea why change point detection in dynamic graphs is
useful, and specifically why one may want to use attributed graphs
rather than graphs without attributes. Finally, the latent process
model is a new model, and it would be beneficial to specify early in
the introduction the relationship between this model and other models
that the reader may be familiar with, namely the mixed-membership
stochastic blockmodel and the latent space models of Hoff et al. To
address these weakness, I would suggest to add a few paragraphs
preceding the current section I.A providing context and relationship
to existing models. The authors should find it useful to refer to
other papers published in the Trans on SP to get an idea on the type
of format, organization and style of writing that the TSP reader
expects.

Furthermore, there are terms and quantities that are introduced
without definition, e.g., the bracket notation "[]", e.g., "[K+1]",
the "cadlag process", and the "rdpm" in Sec 1.A. These are not
notations and quantities that are familiar to a signal processing
audience and should be carefully defined before they are introduced.

Specific comments:
- Page 3, figure 1: Please provide a more descriptive caption with
meanings of the regions and trajectories.
- Page 11, line 38: From the definitions of E and F, they appear to be
scalars, but E+F is defined as the convolution of E and F, which does
not make sense.
- Page 13, lines 10-11: What is the change point t* in this example?
- Page 14, line 49: Change "yields" to "yield"
- Page 15, line 4: A plot of each of the normalized test statistics
T_lambda^l (with the optimal lambda for each statistic) against time
should also be included so the reader can get a sense for how
anomalous this particular time step happens to be.
- Page 17, figure 6: Why are there multiple lines for the second order
approximation and exact model? Are they for different values of r? If
so, the values of r should be indicated somewhere in the plots
themselves (perhaps by reducing the number of values of r shown) or by
listing them in the caption.
- Appendix, page 1, line 31: Change "U-statistics" to "U-statistic"

*** Response

We have reorganized the paper to incorporate the referees comments
about the lack of a proper introduction (Section 2 in the new
manuscript was orignally Section 1 and the new Section 1 now discusses
the motivation behind time-series of attributed graphs). As we allude
to in the manuscript, representations of data as
graphs are ubiquitous in many application
domains and as the graphs representations do change over time,
the notion of a time-series of graphs is natural. Furthermore, in many
application domains, the edges are attributed, e.g., in neuroscience,
the edges can represent either inhibitory or excitatory synapses, in
social networks, the edges can represent the medium of communication
(twitter, email, sms). It is thus desirable for 
any approach that analyzes time-series of
graphs to also extends to attributed graphs
in a transparent manner.

We have also added several paragraphs to Section 2 to clarify/motivate the
latent process model and to compare it with other models such as the
latent vector model or the dot product model. We have also went
through the paper to carefully note/clarify notations for
terms/quantities such as [K+1] or rdpm.

For the specific comments, we addressed them in the paper. In
particular, 
1) the caption for Figure 1 is now updated to contains
description of the regions and tracjectories 
2) It is noted that E and F are meant to be random variables. The
statement that defined E and F had been adjusted accordingly. 
3) t* for the synthetic example is noted to be 11.
4) Figure 7 in the paper now contains  a plot of the
normalized test statistics T_{\lambda}^{l} (with the optimal lambda
for each statistic) against time.
5) The multiple lines for the exact model and the second-order
approximation are for different values of r. The set of vertex process
parameters r considered is also noted in the caption of Figure 5.

### 2012-06-22 ssp2011.tex
Fixed some typo
Revision as submitted to IEEE Trans on SP; Move codes to different repository

\documentclass[10pt,journal,compsoc]{IEEEtran}
\usepackage{fontenc}
\pdfminorversion=4
%\usepackage{lmodern}
\usepackage[utopia]{mathdesign}
\usepackage{graphicx}
%\usepackage{xltxtra}
%\setmainfont[Mapping=tex-text]{Linux Libertine}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{array}
%\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage[tight,normalsize,sf,SF]{subfigure}
\ifCLASSOPTIONcompsoc
\usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
\usepackage[font=footnotesize]{subfig}
\fi
\usepackage{subfig}
\usepackage{parskip}
\usepackage{mathrsfs}
\usepackage{bm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\usepackage[colorlinks=true,pagebackref,linkcolor=magenta]{hyperref}
\usepackage[noadjust]{cite}
\renewcommand\arraystretch{1.2}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\def\mathllap{\mathpalette\mathllapinternal}
\def\mathllapinternal#1#2{%
\llap{$\mathsurround=0pt#1{#2}$}% $
}
\def\clap#1{\hbox to 0pt{\hss#1\hss}}
\def\mathclap{\mathpalette\mathclapinternal}
\def\mathclapinternal#1#2{%
\clap{$\mathsurround=0pt#1{#2}$}%
}
\def\mathrlap{\mathpalette\mathrlapinternal}
\def\mathrlapinternal#1#2{%
\rlap{$\mathsurround=0pt#1{#2}$}% $
}
\bibliographystyle{IEEEtran}
\begin{document}
\title{Attribute fusion in a latent process model for time series of
  graphs}
\author{Minh~Tang, Youngser~Park, Nam~H.~Lee, and Carey~E.~Priebe%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem Minh~Tang, Nam~H.~Lee
 and Carey~E.~Priebe are with the Department of Applied Mathematics
 and Statistics, Johns Hopkins University.\protect \\
  Email: \{mtang10, nhlee, cep\}@jhu.edu
\IEEEcompsocthanksitem Youngser Park is with the Center for Imaging
Science , Johns Hopkins University.\protect \\
  Email: youngser@jhu.edu}%
\thanks{}}
\markboth{IEEE Transactions on Pattern Analysis and Machine
  Intelligence (TPAMI)}{}
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
  Hypothesis testing on time series of attributed graphs has
  applications in diverse areas, e.g., social network analysis
  (wherein vertices represent individual actors or organizations),
  connectome inference (wherein vertices are neurons or brain regions)
  and text processing (wherein vertices represent authors or
  documents).  We consider the problem of anomaly/change point
  detection given the latent process model for time series of graphs
  with categorical attributes on the edges presented in \cite{lee11}.
  Various attributed graph invariants are considered, and their power
  for detection as a function of a linear fusion parameter is
  presented.  Our main result is that inferential performance in
  mathematically tractable first-order and second-order approximation
  models does provide guidance for methodological choices applicable
  to the exact (realistic but intractable) model. Furthermore, to the
  extent that the exact model is realistic, we may tentatively
  conclude that approximation model investigations have some bearing
  on real data applications.
\end{abstract}
\begin{IEEEkeywords}
  Anomaly detection, Attributed Random Graphs, Random Dot
  Product Graphs.
\end{IEEEkeywords}}
\maketitle
\section{Introduction}
\label{sec:introduction}

\subsection{Latent Process Model}
\label{sec:latent-process-model}
The latent process model for time series of attributed graphs was
presented in \cite{lee11}. We summarize here the ideas that are
relevant to our discussion.  Let $G = (V,E)$ be an undirected graph,
equipped with an edge-attribution function $\phi \colon \tbinom{V}{2}
\mapsto \{1,2,\dots, K+1\}$ for some $K$ such that $\phi(e) \leq K$
for $e \in E$ and $\phi(e) \equiv K+1$ for $e \not \in E$. We refer to
$\phi(e)$ as the attribute of edge $e \in E$, and the pair $(G,\phi)$
as an attributed graph. Let $\mathscr{S}$ be the unit simplex in
$\mathbb{R}^{K}$, i.e.,
  \mathscr{S} = \{ \xi \in [0,1]^{K}
  \colon \sum_{k = 1}^{K} \xi_k \leq 1 \}.
A {\em random dot product space} for attributed graphs with vertices
in $[n]$ and edge attributes $[K]$ is a pair $(\mathbf{X},G)$ of
random elements such that
\item $\mathbf{X} = \{X_v\}_{v = 1}^{n}$ is a collection of
  independent $\mathscr{S}$-valued random vectors.
\item $G$ is a random undirected graph with vertex set $[n]$ such that
    \label{eq:1}
    \mathbb{P}(u \sim v \,|\, \mathbf{X}) = \mathbb{P}(u \sim v \, |
    \, X_u,X_v) = \langle x_u, x_v \rangle
  and $\bm{1}(u \sim v \,|\, \mathbf{X})$ and $\bm{1}(u' \sim
  v' \,|\, \mathbf{X})$ are independent whenever $(u,v) \not =
  (u',v')$. Furthermore, we also have 
    \mathbb{P}(\phi(\{u,v\} = k \, | \, u \sim
    v) = x_{u,k} x_{v,k}. 
We say that a c\'{a}dl\'{a}g process $W \colon [0,\infty) \mapsto
[K+1]^{n}$ induces the sequence of random dot product spaces
$\mathscr{V} = (\mathbf{X}(t), G(t))_{t = 1}^{\infty}$ if
\item Each $(\mathbf{X}(t), G(t))$ is a random dot product space with
  vertices $[n]$ and edge attributes $[K]$. Furthermore, for each $v
  \in [n]$, $k \in [K]$ and $t \in \mathbb{N}$, we have
    \label{eq:54}
  X_{v,k}(t)  = \int_{t - 1}^{t}{ \mathbf{1}\{W_v(\omega) = k\}\, d\omega}
  where we denote $W_{v}(\omega)$ as the $v$-th component of $W(\omega)$ and
  $X_{v,k}$ as the $k$-th component of $X_v$.  
\item  For each $t \in \mathbb{N}$, we have
    \label{eq:2}
    \mathbb{P}(G(t) = g \,|\, \mathscr{F}_{\leq t}) = \mathbb{P}(G(t)
    = g \,|\, \mathbf{X}(t))
where $\mathscr{F}_{\leq t}$ is the $\sigma$-field generated by $\{W(s)
  \colon s \leq t\}$.
We will call any pair $(\mathscr{V}, W)$ that satisfies the above
properties a random dot process model. In particular, we are
interested in the pairs $(\mathscr{V}, W)$ possessing the following
properties:

@@

  $\pi^{(v)}(t)$. $W$ is thus a piecewise stationary, continuous-time

@@

      \pi^{(1)}(t) = \dots = \pi^{(m)}(t) = \pi_1 \\

@@

change-point phenomena for $(X(t),G(t))$ at time $t = t^{*}$. We chose $\{1,2,\dots,m\}$
as the set of vertices with a change in the stationary distribution
and intensity matrix at time $t = t^{*} - 1$ for ease of
discussion. Permutation of the vertex labels does not affect our
subsequent analysis. We will refer to $(t^{*}, m, \pi_0, \pi_1,
\mathbf{Q}_0, \mathbf{Q}_1)$ as the change parameters.  Because $W$ is
completely determined by the change parameters, we often choose to
omit $W$ and only mention $\mathscr{V}$ when referring to random dot
process models (rdpm) with change-point phenomena. For a rdpm

@@

  \pi_0, \pi_1, \mathbf{Q}_0, \mathbf{Q}_1)$. The first order approximation
  $\bar{\mathscr{V}}$ of $\mathscr{V}$ is the sequence $\{(\bar{X}(t),
  \bar{G}(t)\}_{t = 1}^{\infty}$ of {\em independent} random dot product
  spaces such that

@@

  \caption{An illustration of the kidney-egg model for $G(t^{*})$.}

@@

position graphs \cite{marchette08:_predic,scheinerman10:_model,hoff02:_laten}.

>>>>

\documentclass[10pt,twocolumn]{IEEEtran}
\usepackage{fontenc}
\pdfminorversion=4
%\usepackage{lmodern}
%\usepackage[utopia]{mathdesign}
\usepackage{graphicx}
%\usepackage{xltxtra}
%\setmainfont[Mapping=tex-text]{Linux Libertine}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{array}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage[tight,normalsize,sf,SF]{subfigure}
\ifCLASSOPTIONcompsoc
\usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
\usepackage[font=footnotesize]{subfig}
\fi
\usepackage{subfig}
\usepackage{parskip}
\usepackage{mathrsfs}
\usepackage{bm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
%\usepackage[colorlinks=true,pagebackref,linkcolor=magenta]{hyperref}
\usepackage[noadjust]{cite}
\renewcommand\arraystretch{1.2}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\def\mathllap{\mathpalette\mathllapinternal}
\def\mathllapinternal#1#2{%
\llap{$\mathsurround=0pt#1{#2}$}% $
}
\def\clap#1{\hbox to 0pt{\hss#1\hss}}
\def\mathclap{\mathpalette\mathclapinternal}
\def\mathclapinternal#1#2{%
\clap{$\mathsurround=0pt#1{#2}$}%
}
\def\mathrlap{\mathpalette\mathrlapinternal}
\def\mathrlapinternal#1#2{%
\rlap{$\mathsurround=0pt#1{#2}$}% $
}
\bibliographystyle{IEEEtran}
\begin{document}
\title{Attribute fusion in a latent process model for time series of
  graphs}
\author{Minh~Tang, Youngser~Park, Nam~H.~Lee, and Carey~E.~Priebe*%
\thanks{Minh~Tang is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (email: mtang10@jhu.edu)}%
\thanks{Youngser~Park is with the Center of Imaging Science, Johns
  Hopkins University, Baltimore, MD 21218
  USA. (email:youngser@jhu.edu)}%
\thanks{Nam~H.~Lee is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (email: nhlee@jhu.edu)}%
\thanks{Carey~E.~Priebe is with the Department of Applied Mathematics and
  Statistics, Johns Hopkins University, Baltimore, MD 21218
  USA. (phone: 410-516-7200; fax:410-516-7459; email: cep@jhu.edu)}}%
%\markboth{IEEE Transactions on Signal Processing}{}
% \begin{IEEEkeywords}
%   Anomaly detection, Attributed Random Graphs, Random Dot
%   Product Graphs.
% \end{IEEEkeywords}}

\maketitle
\begin{abstract}
  Hypothesis testing on time series of attributed graphs has
  applications in diverse areas, e.g., social network analysis
  (wherein vertices represent individual actors or organizations),
  connectome inference (wherein vertices are neurons or brain regions)
  and text processing (wherein vertices represent authors or
  documents).  We consider the problem of anomaly/change point
  detection given the latent process model for time series of graphs
  with categorical attributes on the edges presented in \cite{lee11}.
  Various attributed graph invariants are considered, and their power
  for detection as a function of a linear fusion parameter is
  presented.  Our main result is that inferential performance in
  mathematically tractable first-order and second-order approximation
  models does provide guidance for methodological choices applicable
  to the exact (realistic but intractable) model. Furthermore, to the
  extent that the exact model is realistic, we may tentatively
  conclude that approximation model investigations have some bearing
  on real data applications.
\end{abstract}
\ifCLASSOPTIONpeerreview
\begin{center} \bfseries EDICS Category: SSP-SSAN, SSP-DETC \end{center}
\fi
%\pagebreak
\section{Introduction}
\label{sec:introduction}
The notion of data as a time series is fundamental to signal
processing. However, time-series analysis had been mostly confined to
data in Euclidean spaces, while the representation of data as graphs
with the vertices as the entities and the edges as the relationships
between the entities is now ubiquitous in many application domains,
e.g., social networks (wherein vertices represent individual actors or
organizations), neuroscience (wherein vertices are neurons or brain
regions), and text processing (wherein vertices represent authors or
documents). Furthermore, in many of these application domains, the
edges between the vertices, in addition to representing the presence
of a relationship, can also be use to identify the attributes
associated with that relationship. For example, in social network
graphs where edges represent email communications, the attributes
signify the context of the email messages. Finally, in many of these
application domains, the underlying data changes over time, and thus a
representation of data as a time-series of attributed graphs is
natural. The analysis of time-series of (attributed) graphs thus
presents a natural and challenging extension to the signal processing
community.

We mention here some related work. The notion of signal
processing for graphs appears to be in an emerging
state, with significant research activity in constructing wavelet
representations for graphs, e.g., 
\cite{hammond10:_wavel,r.coifman06:_diffus,crovella03:_graph,narangss:_perfec}. 
The formulation of graph analysis problems in a signal processing framework
is also of potential interest, e.g, \cite{miller10:_towar_euclid}.

One of the main exploitation tasks in time-series analysis is the
problem of anomaly/change-point detection, where anomaly is broadly
interpreted to mean deviation from some ``normal'' pattern and
change-point is the time window at which the anomalous deviation
occurs. In the context of a time-series of graphs, an anomaly can be
for example the presence of a specific subgraph, or frequent occurence
of some subgraphs, or other more vague notions such as existence of a
community structure.

In this paper, we investigate the anomaly/change-point detection
problem for time-series of attributed graphs. In the spirit of
time-series analysis, we approach this problem via test statistics
based on the moving average of some graph invariants. We derive
limiting distributions for these test statistics under a generative
model for the attributed graphs. The experimental evidence indicate that these
test statistics have sufficient power against the null hypothesis for
a general class of change-point problem but that many interesting
questions remain to ponder.

Our paper is structured as follows. We present an overview of the
generative model for time-series of attributed graphs \cite{lee11} in
\S\ref{sec:latent-process-model}. The change-point detection problem
along with the relevant test statistics are described in
\S\ref{sec:change-point-detect}. The distributions of the test
statistics and the power estimates for hypothesis testing using
these test statistics are given in \S\ref{sec:power-estimates}. 
\S\ref{sec:inference-examples} presents a
synthetic data example and an example using the Enron email corpus to
illustrate our methodology.
\section{Latent Process Model}
\label{sec:latent-process-model}
The abundance of graph representations in diverse application domains
leads to a proliferation of random graph models to model the
association among the entities. In addition to the classical
Erd\"{o}s-R\'{e}nyi model \cite{Erdos1959}, there is, the exponential
graph models \cite{holland81}, the stochastic block model
\cite{Holland1983,Wang1987}, the latent position model
\cite{hoff02:_laten}, the dot product model for unattributed graphs
\cite{young07:_random} and the latent process model for attributed
graphs \cite{lee11}. We now describe briefly the latent position model
of \cite{hoff02:_laten} and the dot product model of
\cite{young07:_random} as they are closely related to the latent
process model in \cite{lee11}. The latent process model serves as
the generative model for our attributed graphs and thus plays a
key role in our setup of the anomaly detection problem in 
\S~\ref{sec:change-point-detect}.

The latent space model is one where each vertex is associated with a
latent random vector. There may also be additional covariate
information but those are not relevant to the current exposition. The
vectors are independent and identically distributed and the
probability of an edge between two nodes depends only on their latent
vectors. Conditioned on the latent vectors, the presence of each edge
is an independent Bernoulli trial. For example, if $p_{uv}$ is the
probability of an edge between the vertices $u$ and $v$, then $p_{uv}
= \tfrac{1}{1 + \exp( - \|z_u - z_v\|)}$ where $z_u$ and $z_v$ are the
latent vectors associated with vertex $u$ and $v$, respectively.
Another example of a latent space model is the random dot product
graph model \cite{young07:_random}. Under the random dot product graph
model, the probability of an edge between two vertices is given
by the dot product of their respective latent vectors, i.e., $p_{uv} =
\langle z_u, z_v \rangle$. Parts of the motivation of the random dot
product graph model is that, in a social network with edges
indicating friendships, the components of the vector may be
interpreted as the relative interest of the individual in various
topics. The magnitude of the vector can be interpreted as how
talkative the individual is, with more talkative individuals more
likely to form relationships. Talkative individuals interested in the
same topics are most likely to form relationships while individuals
who do not share interests are unlikely to form relationships.

The latent process model for time-series of attributed graphs was
presented in \cite{lee11}. The main ideas underlying the model is as
follows. The model is motivated by the assumption that each vertex is
associated with a stochastic process, in this case a finite-state
continuous time Markov chain. The stochastic processes then associate
to each vertex $v$, for the time interval $(t-1,t]$, a (latent) random
vector $X_v(t)$ in the unit simplex with the $k$-th entry of $X_v(t)$
denoting the proportion of time the stochastic process spent in state
$k$. The probability of interactions between vertices $u$
and $v$ during $(t-1, t]$ is then given by the dot product of $X_u$
and $X_v$ a la the dot product model of \cite{young07:_random}, with a
slight modification to account for the presence of attributes on the
edges.

We now give the necessary details to make the above ideas precise. Let $G =
(V,E)$ be an undirected graph, equipped with an edge-attribution
function $\phi \colon \tbinom{V}{2} \mapsto [K+1] = \{1,2,\dots, K+1\}$ for
some $K$ such that $\phi(e) \leq K$ for $e \in E$ and $\phi(e) \equiv
K+1$ for $e \not \in E$. We refer to $\phi(e)$ as the attribute of
edge $e \in E$, and the pair $(G,\phi)$ as an attributed graph. Let
$\mathscr{S}$ be the unit simplex in $\mathbb{R}^{K}$, i.e.,
\begin{equation}
  \mathscr{S} = \{ \xi \in [0,1]^{K}
  \colon \sum_{k = 1}^{K} \xi_k \leq 1 \}.
\end{equation}
A {\em random dot product space} for attributed graphs with vertices
in $[n]$ ($[n] = \{1,2,\dots,n\}$) and edge attributes $[K]$ ($[K] =
\{1,2,\dots,K\}$) is a pair $(\mathbf{X},G)$ of
random elements such that
\begin{enumerate}
\item $\mathbf{X} = \{X_v\}_{v = 1}^{n}$ is a collection of
  independent $\mathscr{S}$-valued random vectors.
\item $G$ is a random undirected graph with vertex set $[n]$ such
  that the edges of $G$ are conditionally independent given
  $\mathbf{X}$ and that
  \begin{equation}
    \label{eq:1}
    \mathbb{P}(u \sim v \,|\, \mathbf{X}) = \mathbb{P}(u \sim v \, |
    \, X_u,X_v) = \langle x_u, x_v \rangle.
  \end{equation}
 Furthermore, we also have 
  \begin{equation}
    \label{eq:11}
    \mathbb{P}(\phi(\{u,v\}) = k \, | \, u \sim
    v) = \frac{x_{u,k} x_{v,k}}{\langle x_u, x_v \rangle}
  \end{equation}
  i.e., the pair $\{u,v\}$ is an edge in $G$ with attribute $k$ with
  probability $x_{u,k} x_{v,k}$.  
\end{enumerate}
We say that a c\'{a}dl\'{a}g (right continuous with left limit)
process $W \colon [0,\infty) \mapsto [K+1]^{n}$ induces the sequence
of random dot product spaces $\mathscr{V} = (\mathbf{X}(t), G(t))_{t =
1}^{\infty}$ if
\begin{enumerate}
\item Each $(\mathbf{X}(t), G(t))$ is a random dot product space with
  vertices $[n]$ and edge attributes $[K]$. Furthermore, for each $v
  \in [n]$, $k \in [K]$ and $t \in \mathbb{N}$, we have
  \begin{equation}
    \label{eq:54}
  X_{v,k}(t)  = \int_{t - 1}^{t}{ \mathbf{1}\{W_v(\omega) = k\}\, d\omega}
  \end{equation}
  where we denote $W_{v}(\omega)$ as the $v$-th component of $W(\omega)$ and
  $X_{v,k}$ as the $k$-th component of $X_v$.  
\item  For each $t \in \mathbb{N}$, we have
  \begin{equation}
    \label{eq:2}
    \mathbb{P}(G(t) = g \,|\, \mathscr{F}_{\leq t}) = \mathbb{P}(G(t)
    = g \,|\, \mathbf{X}(t))
  \end{equation}
where $\mathscr{F}_{\leq t}$ is the $\sigma$-field generated by $\{W(s)
  \colon s \leq t\}$.
\end{enumerate}
We will call any pair $(\mathscr{V}, W)$ that satisfies the above
properties a random dot process model.

The random dot process model can be interpreted as follows. During
each time interval $(t-1,t]$, a particular vertex is inclined to
communicate with other vertices on a variety of topics (the topics are
enumerated as the elements of $[K]$). As the inclination of a vertex
$v$ is governed by its underlying stochastic process $W_v$, we can
view the $\{W_v\}$ as encapsulating a notion of changes in the
inclination of the vertices over time. The propensity for
vertex $v$ to communicate about a particular topic $k$ is $X_{v,k}(t)$
and two vertices $u$ and $v$ are more likely to exchange communication
on topic $k$ if the propensity to communicate about topic $k$ is high
for both $u$ and $v$. Eq.~\eqref{eq:1} and Eq.~\eqref{eq:11} attempt
to capture these two assumptions.

In this paper, we are interested in the pairs $(\mathscr{V}, W)$
possessing the following properties:

@@

  $\pi^{(v)}(t)$. $W$ is thus a piece-wise stationary, continuous-time

@@

      \pi^{(1)}((t)) = \dots = \pi^{(m)}(t) = \pi_1 \\

@@

change-point phenomena for $(X(t),G(t))$ at time $t = t^{*}$. We chose
$\{1,2,\dots,m\}$ as the set of vertices with a change in the
stationary distribution and intensity matrix at time $t = t^{*} - 1$
for ease of discussion. Permutation of the vertex labels does not
affect our subsequent analysis. We will refer to $(t^{*}, m, \pi_0,
\pi_1, \mathbf{Q}_0, \mathbf{Q}_1)$ as the change parameters.  Because
$W$ is completely determined by the change parameters, we often choose
to omit $W$ and only mention $\mathscr{V}$ when referring to random
dot process models with change-point phenomena. For a random dot
process model

@@

  Let $\mathscr{V}$ be a random dot process model (rdpm) with change
  parameters $(t^{*}, m, \pi_0, \pi_1, \mathbf{Q}_0,
  \mathbf{Q}_1)$. The first order approximation $\bar{\mathscr{V}}$ of
  $\mathscr{V}$ is the sequence $\{(\bar{X}(t), \bar{G}(t)\}_{t =
    1}^{\infty}$ of {\em independent} random dot product spaces such
  that

@@

  \caption{An illustration of the kidney-egg model for $G(t^{*})$. The
    probability of an edge is $\langle \hat{\pi}_0, \hat{\pi}_0
    \rangle$ for the connections inside the kidney set (black), $\langle
    \hat{\pi}_1, \hat{\pi}_1 \rangle$ for the connections inside the egg set
    (red), and $\langle \hat{\pi}_0, \hat{\pi}_1 \rangle$ for the connections
    between the kidney and egg set.}

@@

position graphs
\cite{marchette08:_predic,scheinerman10:_model,hoff02:_laten}.

Suppose that we have a sequence $\{\mathscr{V}^{r} \colon r > 0 \}$ of
rdpm with vertices $[n]$ and attributes $[K]$ where for each $r > 0$,
$\mathscr{V}^{r}$ has change parameters $(t^{*}, m, \pi_0, \pi_1, r
\mathbf{Q}_0, r \mathbf{Q}_1)$. The parameter $r$ can be thought of as
the vertex process rate, i.e., the waiting time decreases
exponentially as $r$ increases. Let us now consider the sequence of
first approximations $\bar{\mathscr{V}}^{r}$ and second approximations
$\widehat{\mathscr{V}}^{r}$ of $\mathscr{V}^{r}$. We note that
$\bar{\mathscr{V}}^{r_1} = \bar{\mathscr{V}}^{r_2}$ for any $r_1, r_2
> 0$. Let us then denote by $\bar{\mathscr{V}}$ the (a.e.) unique
first order approximation of $\mathscr{V}^{r}$ for $r > 0$. If we
denote by $\widehat{\Sigma}_{0}^{(r)}$ and
$\widehat{\Sigma}_{1}^{(r)}$ the matrices $\widehat{\Sigma}_0$ and
$\widehat{\Sigma}_1$ for $\widehat{\mathscr{V}}^{r}$, then
$\widehat{\Sigma}_0^{(r)} = \widehat{\Sigma}_0^{(1)}/r$ and
$\widehat{\Sigma}_{1}^{(r)} = \widehat{\Sigma}_1^{(1)}/r$. Therefore,
as $r \rightarrow \infty$, $\widehat{\Sigma}_0^{(r)} \rightarrow
\bm{0}$ and $\widehat{\Sigma}_1^{(r)} \rightarrow \bm{0}$ and so
$\widehat{\mathscr{V}}^{r} \overset{\mathrm{d}}{\longrightarrow}
\bar{\mathscr{V}}$. This indicates that for sufficiently large $r$,
there is little statistical difference among the
$\bar{\mathscr{V}}$, $\mathscr{V}^{r}$ and $\widehat{\mathscr{V}}^{r}$
(\cite[Theorem 2]{lee11}). The behavior of
$\widehat{\mathscr{V}}^{r}$ for ``small'' $r$ is less
clear. $\bar{\mathscr{V}}$ will thus serve as the basis for our
subsequent analysis on graph invariants for attributed random
graphs.
